Classical Simulation of Quantum Error Correction in a Fibonacci Anyon Code

### 2016-01-19 fib2.tex
try single column
commit edits
more changes

\appendix

\section*{Supplementary Material}

The goal here is to describe the algorithm for
computing charge contained regions built from 
arbitrary collection of tiles.
To do this we first describe our approach
to topological quantum field theory,
then we discuss the data structure 
(combinatorial curve diagrams) and the algorithm
that operates on these structures.

%We give an informal introduction to
%anyonic topological quantum
%field theory in $(2+1)$-dimensions.
%See \cite{beverland2014} for a similar treatment.
%Note that these theories give rise to
%\emph{doubled} anyonic charges, and so
%the Fibonacci example in the main text simulates
%one half of such a system. % MEH!

%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%

\section{Modular Functors}

As in algebraic topology, we wish to extract
algebraic information about topological objects.
This can be viewed as a kind of representation theory
of topological objects: homeomorphisms of surfaces.

We list the axioms for a 
\emph{2-dimensional unitary topological modular functor}.
 % abbreviated as \emph{MF}.
There are quite alot of details that we leave out,
%For the details of this construction, see
see
\cite{Turaev1994}, \cite{Walker1991} or \cite{Bakalov2000}.
For a more pedestrian treatment, see \cite{Freedman2002simulation} or \cite{Beverland2014}.

\newcommand{\F}{\mathscr{H}} % H or F ?
\newcommand{\C}{\mathbb{C}}
\newcommand{\A}{\mathcal{A}}

For our purposes a \emph{surface} will be an
oriented $2$-dimensional manifold with boundary.
We will not require surfaces to be connected.
%Each boundary component is oriented (either explicitly or
%otherwise by using the orientation of the manifold)
Each boundary component will inherit the manifold orientation,
contain a distinguished \emph{base point},
and be labelled with
%an element from a finite set $\mathbb{A}=\{1, a, b, c ...\}.$
an element from a finite set $\A.$
This is the set of ``anyon labels'', and comes
equiped with a vacuum element $\vac$ 
and an involution $\ \widehat{}\ $ such that $\widehat{\vac}=\vac.$

We will mostly be concerned with \emph{planar} such
surfaces, ie. disc(s) with zero or more holes.

A modular functor $\F$ associates to every
surface $M$ a finite dimensional complex
vector space $\F(M)$, called the \emph{fusion space} of $M$.
A map of surfaces $f:M\to N$ will be understood to be
a topological homeomorphism that preserves boundary
labels, orientation, and base points.
The modular functor associates such an $f$ to a
unitary transformation $\F(f) : \F(M)\to\F(N)$
that only depends on the isotopy class of $f.$
Functoriality requires that $\F$
respect composition of homeomorphisms.

We have the following axioms for $\F.$

{\bf Axiom I.}
The fusion space of an empty surface is one dimensional, 
$\F(\phi) \cong \C.$
For $M_a$ a disc with boundary label $a$ we have
$\F(M_\vac)\cong \C$ and 
$\F(M_a)\cong 0$ for $a\ne \vac.$
For annulus $M_{a,b}$ with boundary labels
$a, b$ we have
$\F(M_{a,\widehat{a}})\cong \C$ and 
$\F(M_{a,b})\cong 0$ for $a\ne \widehat{b}.$

{\bf Axiom II.}
The disjoint union of two surfaces $M$ and $N$ 
is associated with
the tensor product of fusion spaces:
$$
    \F(M\amalg N) \cong \F(M)\otimes \F(N).
$$
This is natural from the point of view of quantum
physics, where the Hilbert space of two disjoint
systems is the tensor product of the space for
each system.

{\bf Axiom III.}
%Gluing.
Denote a surface $M$ with (at least) two boundary
components labelled $a, b$ as $M_{a,b}$. 
Let $N$ be the surface obtained from $M$ by gluing
these two boundary components together.
Glueing must respect the orientation of the two boundary
components as well as identifying their base points.
If we constrain $b=\widehat{a}$ then the fusion space
of $M_{a,\widehat{a}}$ embedds unitarily in the fusion
space of $N.$ Moreover, there is an isomorphism called
a \emph{gluing map}:
$$
    \bigoplus_{a\in\A} \F(M_{a,\widehat{a}}) \xrightarrow{\cong} \F(N).
$$
The seam along which gluing occurs can be associated
with an observable as follows.
We take a gluing map $g$. Projectors $P_a$ onto the
summands in the above direct sum:
$$
    P_a: \bigoplus_{b\in\A} \F(M_{b,\widehat{b}}) \to \F(M_{a,\widehat{a}}).
$$
then the observable will be the set of operators $\{P_a g^{-1}\}_{a\in\A}.$

A common operation is to glue two separate surfaces.
We can do this by first taking disjoint union (tensoring
the fusion spaces)
and then gluing.
Here we show this process applied to two surfaces $M$ and $N$. 
\begin{center}
\includegraphics[]{pic-glue.pdf}
\end{center}
We display the surface $N$ with the $\widehat{a}$
boundary on the outside, 
to show more clearly how $N$ fits into $M$.
In the glued manifold we
indicate the placement of $M$ and $N$ and the seam
along which glueing occured,
as well as the identification of base points.

%\emph{XXX show orientation of boundaries ?}
%\emph{XXX displaymath direct sum}
%\emph{XXX $N...$ show ellipses?}

A most important linear map, called an $F-$move,
is constructed from two applications of a
gluing map as follows.

\begin{widetext}
\begin{center}
\includegraphics[]{pic-glue-fmove.pdf}
\end{center}
\end{widetext}

Here we have shown a manifold $M_{abc}$ with 
three labelled boundaries, as well as two separate
ways of gluing pairs-of-pants to get $M.$

As we have seen, an oriented simple closed path 
 contained in the interior of $M$ can serve as
the seam along which gluing occurs and so we call
these paths \emph{observables}.
%Any oriented simple closed path contained in the interior of $M$ we will
%call an \emph{observable}.
Such a path that encloses two boundary components 
we call a \emph{2-observable}.
\emph{(what does enclose mean exactly?)}
A function $c:[0,1]\to M$ such that $c(0)\in \partial M$ and $c(1)\in \partial M$
are two separate components of the boundary we call a \emph{2-curve.}
The following important fact will be used extensively
in the combinatorial formulation of a modular functor below:
up to isotopy (not preserving $\partial M$),
there is a (natural) bijection between
$2$-observables and $2$-curves.
\emph{(restrict to $M$ a disc with at least two holes?)}
Here we show two such instances of this
bijection:
\begin{center}
\includegraphics[]{pic-2-curve.pdf}
\end{center}
The bijections works as follows.
To construct a $2$-observable from a $2$-curve $c$ that
joins two boundary components $M_1$ and $M_2$,
take a small neighbourhood of
the union $c([0,1])\cup M_1 \cup M_2.$
This neighbourhood will have boundary a simple closed
curve that encloses $M_1$ and $M_2.$
\emph{(check the orientation of the $2$-observable has the correct interior)}
\emph{(the $2$-observable forgets the direction of the $2$-curve)}
Going the other direction...

The end points of $2$-curves are free to terminate anywhere
on the boundary component, but we can fix the endpoints using
the base points.
A simple way to do this is to just require the
$2$-curve to begin and end on base points.

By a \emph{hole} of $M$ we mean a component of $\partial M$.

A single $F$-move is seen to relate a pair of $2$-curves
that share a hole.
Continuing in this way, we can chain these $2$-curves together:
\begin{center}
\includegraphics[]{pic-chain.pdf}
\end{center}

%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%

%\section{Interaction picture}
\section{Topological quantum field theory}

In two spacial dimensions, particle exchange is
captured by the braid group.
World lines of such particles 
become tangled (braided) in $(2+1)$-dimensions.
For example, three particle statistics is captured
by the braid group on three strands $B_3.$
This group is generated by elements $\sigma_1$
and $\sigma_2$ which we represent pictorially as:
\begin{center}
\includegraphics[]{pic-braid-group.pdf}
\end{center}

These pictures make clear how the group
multiplication works, for example:
\begin{center}
\includegraphics[]{pic-braid-group-1.pdf}
\end{center}
\begin{center}
\includegraphics[]{pic-braid-relation.pdf}
\end{center}

%Braid group acts (naturally) on states.

These diagrams are to be understood up to
continuous deformation of the ambient space.
That is, we allow the ambient $(2+1)$-dimensional
space with these worldlines subtracted to be continuously
deformed.
At the top and bottom of this space is 
a $2$-dimensional space (a disc) with holes in it.
Such a closed disc with $n$ holes we denote as $D_n.$
So this is how we arrive at modelling the system as
a disc with holes in it.
The observables of the system measure charge within
a region bounded by a directed closed curve.
\begin{center}
\includegraphics[]{pic-observable.pdf}
\end{center}

The defining feature of an Abelian anyon theory
is that the state is completely determined once
we measure the charge of each hole.
In the language of the toric code (homology),
logical operators (1-chains)
cancel where they overlap
with opposite orientation:
\begin{center}
\includegraphics[width=0.7\columnwidth]{pic-abelian.pdf}
\end{center}

This is not at all the case in general;
for non-Abelian theories there can be extra
degrees of freedom associated with these
observables. For example, in
a system supporting Fibonacci anyons,
we can have two Fibonacci anyons with total
charge either vacuum or Fibonacci:
\begin{center}
\includegraphics[]{pic-fusion.pdf}
\end{center}
so the state space of two fibonacci anyons
is $2$-dimensional.
Such a space associated with observing total charge
of multiple anyons is called the \emph{fusion} space
of those charges.

These observables satisfy various relations,
but in particular, non-overlapping observables commute.
So if we nest a maximal set of these observables together
we can pin-down a basis for the fusion space.
Such a nesting is tree-like, and if we confuse the distinction
between holes and observables, this gives rise to the
fusion tree notation.
\begin{center}
\includegraphics[]{pic-tree-0.pdf}\includegraphics[]{pic-tree-1.pdf}
\end{center}

Given a disc with $n$ holes, we define the \emph{mapping
class group} $MCG(D_n)$ as the group of topological
symmetries (homeomorphisms) of the space $D_n,$
modulo continuous deformations (isotopy).
Such symmetries we require to fix the boundary
of the disc, but not necessarily to fix the holes.
Elements $f\in MCG(D_n)$ will also act on anything that
resides in $D_n.$ 
In particular, if we draw a line
connecting the holes:
\begin{center}
\includegraphics[]{pic-twist.pdf}
\end{center}

Such an embedding $c : [0, 1] \to D$ with $c(0), c(1) \in \partial D$
and touching all holes in $D_n$ we call a \emph{curve diagram}~\cite{Dehornoy2002},
(or just a \emph{curve} when the context is clear).

We can think of braids acting on states in a Schrodinger picture,
or alternatively, elements of the mapping class group acting on
observables in a Heisenberg picture:
\begin{center}
\includegraphics[]{pic-interaction.pdf}
\end{center}

We would hope that the groups $B_n$ and $MCG(D_n)$ are
isomorphic, and this is indeed the case~\cite{Kassel2010}.
One can see this as follows. Dragging the space $D_n$ along
a braid creates a deformation, the end point of which is a
homeomorphism in $MCG(D_n).$ Going the other direction,
given $f\in MCG(D_n)$ and a curve diagram $c,$ we think of
pulling the ends of $f\circ c$, thereby moving (braiding) the holes,
until we are back at the original curve $c.$

%This discussion makes it clear we have picked out a
%specific curve $c$ that we use as a reference point for
%other curves.

The importance of this isomorphism: the simulation
algorithm calculates outcomes of observations. This
boils down to a change of basis, to one that includes
the required observable.
$F$-moves allow us to re-associate observables along
a curve diagram. The mapping class group acts
transitively on curve diagrams (\emph{why??}).

%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%

\section{Combinatorial curve diagrams}

The basic data structure involved in the
simulation we term a \emph{combinatorial curve
diagram.}
% XXX define "piece of curve"
Firstly, we will require each curve to intersect 
the edges of tiles transversally,
and in particular a curve will not touch a tile corner.

For each tile in the lattice,
we store a combinatorial
description of the curve(s) intersected with that tile.
Each component of such an intersection we call a \emph{piece-of-curve.}
\begin{center}
\includegraphics[]{pic-cells.pdf}
\end{center}

We follow essentially the same approach as taken in \cite{Abramsky2007} 
to describe elements of a Temperley-Leib algebra, but
with some extra decorations.
The key idea is to store a \emph{word} for each tile, comprised of
the letters $\bigl<$ and $\bigr>$.
%The encoding works as follows.
Reading in a clockwise direction around the edge of
the tile from the top-left corner,
we record our encounters with each piece-of-curve,
writing~$\bigl<$ for the first encounter, and~$\bigr>$ for the
second.
We may also encounter a dangling piece-of-curve
(the head or the tail), so we use another symbol $*$ for this.
The words for the above two tiles will then be 
$\bigl<\bigl<\bigr>\bigr>\bigl<\bigr>$ and $\bigl<\bigr>*\bigl<\bigl<\bigr>\bigr>.$
When the brackets are balanced,
each such word will correspond one-to-one with an intersection
of a curve in a tile, up to a continuous deformation of the interior of the tile.
Ie. the data structure 
will be insensitive to any continuous deformation of the interior of the tile,
but the simulation does not need to track any of these degrees of freedom.

\begin{center}
\includegraphics[]{pic-cells-0.pdf}
\end{center}

We will also need to record
various other attributes of these curves,
and to do this we make this notation more elaborate
in the paragraphs {\bf (I)}, {\bf(II)} and {\bf(III)} below.
Each symbol in the word describes an intersection of
the curve with the tile boundary,
and so as we decorate these symbols these decorations will
apply to such intersection points.

\begin{center}
\includegraphics[]{pic-cells-1.pdf}
\end{center}

{\bf (I)} We will record the direction of each piece-of-curve,
this will be either an {\tt in} or {\tt out} decoration for each symbol.
Such decorations need to balance according to the brackets.
The decorated symbols $*_{\mbox{\tt in}}$ and 
$*_{\mbox{\tt out}}$ 
will denote respectively either
%the head, $c(1)$ or the tail $c(0)$ of a curve.
the head or the tail of a curve.
The words for the diagrams above now read as
$ \bigl<_{\mbox{\tt in}}\bigl<_{\mbox{\tt out}}\bigr>_{\mbox{\tt in}}
    \bigr>_{\mbox{\tt out}}\bigl<_{\mbox{\tt out}}\bigr>_{\mbox{\tt in}}$
and
$ \bigl<_{\mbox{\tt in}}\bigl>_{\mbox{\tt out}}*_{\mbox{\tt in}}
    \bigr<_{\mbox{\tt out}}
    \bigr<_{\mbox{\tt in}}\bigl>_{\mbox{\tt out}}\bigr>_{\mbox{\tt in}}.
$

{\bf (II)} We will record,
for each intersection with the tile edge, 
a numeral indicating which of the four
sides of the tile the
intersection occurs on.
Numbering these clockwise from the top as $1, 2, 3, 4$ we have for the above curves: 
$\bigl<_1\bigl<_2\bigr>_2\bigr>_3\bigl<_3\bigr>_4$ 
and $\bigl<_1\bigr>_1*_2\bigl<_3\bigl<_3\bigr>_4\bigr>_4.$

{\bf (III)} Finally, we will also decorate these symbols with anyons.
This will be an index to a leaf of a (sum of) fusion tree(s).
This means that anyons only reside on the curve close
to the tile boundary,
and so we cannot have more than two anyons
for each piece-of-curve. 
The number of such pieces is arbitrary, and so this
is no restriction on generality.

\begin{center}
\includegraphics[]{pic-cells-2.pdf}
\end{center}

In joining tiles together to make a tiling we will
require adjacent tiles to agree on their shared boundary.
This will entail sequentially pairing symbols in the
words for adjacent tiles
and requiring that 
the {\tt in} and {\tt out} decorations are matched.
Because the word for a tile proceeds conter-clockwise
around the tile, this pairing will always reverse the
sequential order of the symbols of adjacent tiles.
For example, given the above two tiles we sequentially pair the 
$\bigl<_{\mbox{\tt out},2}\bigr>_{\mbox{\tt in},2}$ 
and $\bigr>_{\mbox{\tt out},4}\bigr>_{\mbox{\tt in},4}$
symbols with opposite order so that
$\bigl<_{\mbox{\tt out},2}\sim\bigr>_{\mbox{\tt in},4}$
and $\bigr>_{\mbox{\tt in},2} \sim \bigr>_{\mbox{\tt out},4}.$

%Two other consistency relations are enforced on such a combinatorial curve diagram:
%we require adjacent tiles to agree on their boundaries, 
%and every curve diagram
%must have two ends.
%One final consistency
%relation is enforced by requiring 
%We require every curve diagram to have two ends, this means
%that there are no loops.

Note that in general this data structure will store many disjoint curve diagrams
$c_i:[0,1]\to D_{n_i}$ within a disc $D_m$ where $\sum n_i = m.$

%For a given piece-of-curve, we can subtract the numeral with
%the {\tt in} label from the numeral with the {\tt out} label
%to get an integer in $\{ \}$ WRONG
%We can also just use the numerals of the boundary edges
%that the piece-of-curve intersects with, subtracting the
%``exit'' boundary numeral from the ``entry'' boundary numeral. WRONG

For each piece-of-curve, apart from a head or tail, there is an associated 
number we call the \emph{turn number}. This counts the number
of ``right-hand turns'' the piece-of-curve makes as it
traverses the tile, with a ``left-hand turn'' counting as $-1.$
(To be more rigorous, we would define this number using the
winding number of the simple closed curve formed by the
piece-of-curve adjoined to a segment of the boundary of the tile 
traversed in a clockwise direction.)
This number will take one of the values $-2, -1, 0, 1, 2:$
\begin{center}
\includegraphics[]{pic-cells-3.pdf}
\end{center}

%Two disjoint curves can be joined by...
%
%Curves can be simplified along sections without any anyons, ...

% Each piece will have +1, +2, 0, -1, -2 right-hand turns...

% mention Jones' planar algebras ?

%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%

\section{The paperclip algorithm}

Anyons are transported around the lattice
by moving them along tile edges.
In general, such a transport will intersect with a
curve diagram in many places.
Each such intersection is transverse,
and we use each intersection point to cut
the entire transport into smaller paths each of
which touch the curve diagram twice.
%Each intersection with a curve diagram
%will then be transverse, and we
%decompose the entire path into a sequence of
%paths each of which 
%join consecutive intersections.
%Transport of an anyon can be decomposed into
%moves between adjacent components of a curve
%diagram.
The origin and destination of such an anyon path
now splits the curve diagram $c:[0, 1]\to D_n$ 
into three disjoint pieces which we term
\emph{head}, \emph{body} and \emph{tail}, where
the head contains the point $c(1)$, the tail
contains $c(0)$ and the body is the third piece.
These arise with various arrangements, but here
we focus on one instructive case, the
other cases are similar:
%We look at the particular case of moving along
%one edge of a tile,
transporting along one edge of a tile \emph{forwards} 
(from tail to head) along a curve diagram:
\begin{center}
\includegraphics[]{pic-move-anyon.pdf}
\end{center}

This arrangement is equivalent (homotopic) to one of four 
``paperclips'', which we distinguish between by counting how
many \emph{right-hand turns} are made along the body of the curve diagram.
We also show an equivalent (homotopic) picture where the
curve diagram has been straightened, and the resulting distortion
in the anyon path:
\begin{center}
\includegraphics[]{pic-paperclip.pdf}
\end{center}
The sequence of anyons along the head, body and tail, we denote as $H, B$ and $T,$
respectively.
These sequences have the same order as the underlying curve diagram, and 
we use
$H^r, B^r$ and $T^r$ to denote the same anyons with the reversed order.
Using the above diagram, we can now read off the $R$-moves for each
of the four paperclips:
\begin{align*}
-2:&\ R[B] \\
-4:&\ R[H^r]\ R[H]\ R[B] \\
+4:&\ R[B]\ R[T]\ R[T^r] \\
+6:&\ R[H^r]\ R[H]\ R[B]\ R[T]\ R[T^r] \\
\end{align*}
where notation such as $R[B]$ is understood as sequentially clockwise braiding around
each anyon in $B$.

That these four paperclips exhaust all possibilities can be seen by
considering the winding number of the simple closed curve made
by combining the body of the curve diagram with the path followed by
the anyon (appropriately reversing direction as needed).

\section{Computation of homologically non-trivial operators}\label{s:homnontrivial}

Specializing to the Fibonacci case,
we write the non-trivial $F$-moves as the following
skein relations:
\begin{align*}
\includegraphics[]{pic-skein1.pdf}
\end{align*}

The sollid lines represent Fibonacci world-lines.
The dotted lines represent vacuum charges,
and we are free to include these lines or not.
We leave these anyon paths
as undirected because Fibonacci anyons are
self-inverse.
The non-trivial $R$-moves are:
\begin{align*}
\includegraphics[]{pic-skein2.pdf}
\end{align*}

Removing bubbles:
\begin{align*}
\includegraphics[]{pic-bubble.pdf}
\end{align*}

Here we show a process where a 
Fibonacci anyon travels around the torus and
anihilates itself. Twice.
The vertical lines represent a periodic
identification.
\begin{align*}
\includegraphics[]{pic-logops.pdf}
\end{align*}

The state is not normalized.
Also involves post-selection, as there is
another process that involves leakage...
The first equation is an $F$-move, 
the second equation is a translation in the
horizontal direction, and the last equation
follows from the rule for collapsing bubbles.

Continuing in this way, we compute the $k$-fold
logical operator:
\begin{align*}
\includegraphics[]{pic-kfold.pdf}
\end{align*}

where $f_k$ is the $k$-th element of the Fibonacci
sequence $\{1, 1, 2, 3...\}.$

%------------------------------------------------------------------------------------------------------------%
\bibliography{refs2}

>>>>

\bibliography{refs2}
